Noisy Work: A Review of The Uncertainty Visualisation Literature

### 2024-02-05 paper.qmd
finished general uncertainty communication bit

Most research into general uncertainty communication seems to focus specifically on how we communicate risks. Risk and uncertainty are not perfect synonyms of one another. @Spiegelhalter2017 makes the distinction clear by specifying *risk* as *a precise random event* (such as a coin flip) while uncertainty is imprecision (such as the estimated probability associated with a coin that might be biased). For this reason, risk is often seen as a specific case of uncertainty and that well defined sub area has a lot of detailled research. Some markers of probability that have common uses, probabilities such as 0 (impossible), 0.5 (a fair coin flip), 1 (certain) are easy for people to have an internal sense of [@Spiegelhalter2017]. It is incredibly hard to communicate small risks, because people cannot differentiate between 1/100 an 1/1000, and communicating low probability but high impact risks are particularly challenging [@Spiegelhalter2017]. This issue can be somewhat alleviated by communicating the likelihood of an event in relation to another event on a similar scale (such as being struck by lightning) [@Spiegelhalter2017]. The framing of how we communicate risks matters due to the affect heuristic, so there will be a difference in peoples behaviours if, for example, you tell people they have a 90% chance of survival vs a 10% chance of death, and when expressing a frequency, a larger numerator communicates a larger risk [@Spiegelhalter2017]. The reference class of a particular risk should be explicitly stated to avoid confusion, for example if we are communicating the chance of rain, the time period over which we are expecting this chance of rain (i.e. 50% chance today or this week) needs to be specified [@Spiegelhalter2017]. The problem with a risk focus is that it is only a small piece of the uncertainty pie, and very few of these findings can be reasonably assumed to go beyond risk communication. Even less can be assumed to translate to uncertainty visualisation.

@@

Risk and uncertainty are quite different in practice. Uncertainty information is typically included in some form of statistical communication to offer some transparency to the viewer. Data analysis is a long and complicated process, so much so that even experts in the field can find themselves lost with a single incorrect assumption, so uncertainty can help pull back the curtain on this process and prevent people from understanding it as a human agnostic, entirely deterministic process. When science authors discuss "risk avoidance" in the uncertainty communication conversations, it holds with it some implication that heuristics such as risk avoidance are relevant in every conversation about uncertainty communication, however that is not true.

There are some elements of risk communication that seem to translate to uncertaitny communication more broadly. Using words to communicate risk (as well as uncertainty more broadly) is discouraged because people can misinterpret the actual risk involved, so when possible numerical estimates of risk should be communicated, however this can make information harder to understand for those with poor numeracy skills [@Spiegelhalter2017].

likely due uncertainty and risk being near synonymous, it is difficult to read a paper that discusses uncertainty that does not make a nod towards human attitudes towards risk. Even when the uncertainty being discussed in that particular paper has very little to do with risk.

Risk-aversion is the tendancy of people to prefer outcomes of low uncertainty to those outcomes with high uncertainty despute the outcomes with high uncertainty having a higher or equal expected outcome. The story of uncertainty communication is littered with conflated ideas, and that issue is highly visible here. Despite being mentioned in a large proportion of papers that discuss uncertainty risk aversion, seems to have very little to do communicating uncertainty. As a matter of fact, it has so little to do with uncertainty communication that we would not even mention it in this paper were it not for the fact that ignoring it would suggest we had missed an important section of the literature.

 Additionally, risk avoidance is observed when people have a choice between a *deterministic* outcome and a *risky* event that has higher expected value. How this psychological heuristic is therefore related to all uncertainty visualisation despite uncertainty communication being different in both the scenario (i.e. it is highly unlikely that we need to communicate a choice between a perfectly deterministic and an uncertain event) and in the type of uncertainty (since risk and uncertainty are not synonyms).

The regular conflation of risk and uncertainty despite clear and specific differences, without any explanation of how these terms are related, implies that a large portion of uncertainty literature is simply reporting information from any research that is uncertainty adjacent without considerations for the specifics of how these concepts would be related to uncertainty.

It is not even clear that this behaviour is entirely irrational either, considering risk aversion is, at its core, a trade off between bias and variance. If risk-aversion is truly "irrational" behaviour, every statistical textbook that discusses the use of a biased but consistent estimator in a fite sample case (a common example would be specific cases of the maximum likelihood estimator) should immediately toss those chapters out and the authors should feel embarassed for that *clearly* irrational line of thinking. Of course I am making a joke but this does highlight a clear problem in the way risk-aversion is discussed in the literature. Discussing risk-aversion as though it is some diversion from a perfectly rational, mathematical choice, implies there is *no value to a decrease in variance* which is not even true *within mathematics*. The concept of risk aversion treats risk as "noise", as though it is something that only exists to be ignored, but if the common place use of biased MLE's is an indication of anything, it is that *ignoring uncertainty* is the irrational choice.

Selecting a biased is technically statisticians doing "risk aversion", but when used to ironically when it is used to select models it is not viewed as an irrational human fault.

This issue, where psychological considerations seem to be entirely devorced from what is *actually* being discussed extends to the conversation around trust as well. The idea of measuring or being concerned about audiences perception of trust, without first establishing if what we are communicating is *trustworthy*, . The goal of uncertainty visualisation should not be to be trustworthy, but rather to be transparent (which can facilitate trust).

I dont think this is worth a section but it is mentioned a lot in .
- problem of communicating to a group thing (avoid communicating inaccurate information)
- how people feel about uncertainty is irrelevant
  - natural disasters make people uncomfortable but to include it in a conversation about whether or not we should express it is rediculous.
- Numerical risks: Use absolute risks (but relative risk when catastrophic event), for single unique event use percent chance, express change as a proportion, frequency, or percentage (and be clear about reference class), avoid framing bias by showing risk with and without the outcome, keep the denominator fixed when making comparisons, explicit about time interval, For more knowledgeable audiences consider providing epistemic uncertainty and confidence in analysis @Spiegelhalter2017

- General issues when communicating risks: clear objectives, segment audience into target groups, use plain language and limit info to only what is necessary, allow for different levels of interest knowledge and numeracy (a gist level, numerical level, then evidence and uncertainty) @Spiegelhalter2017

Uncertainty visualisation is related to uncertainty communication in the same way that visualisation more broadly is related communication more broadly. Visualisation is fast, it can communicate more complicated ideas that what can be done with text alone, and 
# 4 LITERATURE SUMMARY
- methods used in uncertainty visualisation evaluation remain ad hoc. studies approach issue from a usability perspective instead of asking WHY representations DO or DO NOT work. Studies do not follow any methodology commonly agreed upon. @Kinkeldey2014
- Lack of formalisation and eigour in empirical methdos is an issue that is much broader than uncertainty visualisation (extends to info-vis and related domains). Therefore this paper is a starting point, not a definitive summary @Kinkeldey2014
- The question in these papers is often "does method a work better than method B" which is an engineering approach, where the toal is to improve a particular product rather than create general principals @Kinkeldey2014
- Authors then try to generalise beyond the specific constraints of the test even though the framing does not allow for it @Kinkeldey2014
- Alternatively, we should ask how/why does method A work better than method B, which grounds work in perceptual and cognative theory and provides a framwork @Kinkeldey2014

>>>>

Most research into general uncertainty communication seems to focus specifically on how we communicate risks. Risk and uncertainty are not perfect synonyms of one another. @Spiegelhalter2017 makes the distinction clear by specifying *risk* as *a precise random event* (such as a coin flip) while *uncertainty* (also sometimes referred to as abiguity) is imprecision (such as the estimated probability associated with a coin that might be biased). For this reason, risk is often seen as a specific case of uncertainty and that well defined sub area has a lot of detailled research. Some markers of probability that have common uses, probabilities such as 0 (impossible), 0.5 (a fair coin flip), 1 (certain) are easy for people to have an internal sense of [@Spiegelhalter2017]. It is incredibly hard to communicate small risks, because people cannot differentiate between 1/100 an 1/1000, and communicating low probability but high impact risks are particularly challenging [@Spiegelhalter2017]. This issue can be somewhat alleviated by communicating the likelihood of an event in relation to another event on a similar scale (such as being struck by lightning) [@Spiegelhalter2017]. The framing of how we communicate risks matters due to the affect heuristic, so there will be a difference in peoples behaviours if, for example, you tell people they have a 90% chance of survival vs a 10% chance of death, and when expressing a frequency, a larger numerator communicates a larger risk [@Spiegelhalter2017]. The reference class of a particular risk should be explicitly stated to avoid confusion, for example if we are communicating the chance of rain, the time period over which we are expecting this chance of rain (i.e. 50% chance today or this week) needs to be specified [@Spiegelhalter2017]. The problem with a risk focus is that it is only a small piece of the uncertainty pie, and very few of these findings can be reasonably assumed to go beyond risk communication. Even less can be assumed to translate to uncertainty visualisation.

@@

Risk and uncertainty are not only different in how they are communicated, but also why they are communicated. Risk communication focuses on communicating probability around events. These events are typically a binary event that is known to be random, so a version of communication that does not mention risks (e.g. telling someone they will or will not get cancer instead of communicating their risks about it) inherently carries with it a lack of transparency. Additionally risk communication is often communicated to avoid or informa about an unfavorable outcome, so framing can be important if we want to influence decisions in a certainty way. These motivations do not extend to the communication of uncertainty information more broadly. While uncertainty information *is* often used to improve decisions, it seems the main motivation in its communication is to offer transparency to the viewer about a statistical analysis. That is, it acts as a form of statistical "hedging" for signals found in an analysis. Interviews with experts in statistic back up this primary motivation, as ignoring uncertainty information of often expressed as being similar to fraud or lying and the goal of "improving decisions" is only seen as a secondary outcome of an appropriately hedged signal [@Hullman2020a; @Manski2020]. This concept of uncertainty means that the work on communicating risk does not cover the field of uncertainty communication as a whole. It covers the specific case where the signal itself *is* uncertain, but this work does not naturally extend to the goal of hedging a signal (i.e. when we want to communicate uncertainty as noise). Unfortunately understanding the relevant noise seems to be its own problem, as statisticians and visualisation authors seem to be unaware or confused in how to calculate the uncertainty related to an analysis [@Hullman2020a]. Additionally, communicating uncertainty such that it conveys the appropriate singal supression involves combining uncertainty from multiple different sources, something that is considered to be difficult by even experts in uncertainty communication [@Spiegelhalter2017].

There are some elements of risk communication that seem to translate to uncertaitny communication more broadly. Using words to communicate risk is discouraged because people can misinterpret the actual risk involved, so when possible numerical estimates of risk should be communicated, however this can make information harder to understand for those with poor numeracy skills [@Spiegelhalter2017]. People performing better and preferring with numerical estimates of risk translates to uncertainty communication more broadly *inc (citations for this)*. Risk communication also needs to have clear objectives, use plain language, limit information to only what is necessary, and segment the audience to allow for differences in interest and knowledge [@Spiegelhalter2017]. These general concepts of communication also extend to communication of uncertainty. On the other hand, it is not sensible to assume that probability specific concepts such as framing, intuitiveness of probabilities, or reference classes are still relevant when we consider uncertainty communication as signal supression, not as a synonym of risk or probability communication. In this same vein, most uncertainty communication papers very often discuss risk-aversion as something that needs to be considered despite it being unclear exactly how risk-aversion is related to uncertainty communication *(citations for the papers that do this)*.

In the same sense that risk and uncertainty are different, risk-aversion and amiguity-aversion (which can be considered the uncertainty version of risk aversion) are also not the same. Risk-aversion is the tendency of people to prefer outcomes of low uncertainty to those outcomes with high uncertainty despite the outcomes with high uncertainty having a higher or equal expected outcome. In the case of communicating risk, the only reason this heuristic would be relevant was if risk-aversion was a mistake that needed to somehow be corrected for, rather than a heuristic that *reveals the utility of decreased uncertainty*. Risk-aversion is often treated as a mistake or something that should be avoided by the uncertainty visualisation literature even though that is not necessarily true, depending on what is causing the risk-aversion (e.g. is it due to poorer estimates or increased awareness of negative outcomes). If a particular communication method leads people to place a greater value on certainty or make worse estimates, that is the aspect of importance, not risk-aversion which conflates these factors. The work that advocates for transparency in risk-communication gives you whiplash when it's subtext argues that transparency leads to "incorrect" conclusions. Risk-aversion has a large number of confounding factors that make it difficult to understand if it is something to be considered in risk communication, ambiguity-aversion has a similar problem. Ambiguity-aversion is the tendency of people to prefer to take on certain risks rather than unknown risks. It is essentially risk-aversion applied to risk. For this reason discussions of ambiguity-aversion have the same issues as the conversations about risk-aversion, where they do not consider ambiguity-aversion to be an estimate of the utility of decreased uncertainty, it contains a subtext that argues against transparency, and it is considered a valuable insight in of itself despite conflating many factors.

The idea that risk and ambiguity aversion reveal the utility of decreased variance seems to seldom be considered in the literature, therefore it should be mentioned here, otherwise this issue might continue. However, considering risk aversion is at its core, a trade off between bias and variance, it is bizzare to assume it to be a mistake when statisticians perform these trade offs all the time. If risk-aversion is truly "irrational" behaviour, every statistical textbook that discusses the use of a biased but consistent estimator in a finite sample case (a common example would be specific cases of the maximum likelihood estimator) should be tossed out. Discussing risk and ambiguity aversion as though it is some diversion from a perfectly rational, mathematical choice, implies there is *no value to a decrease in variance which is not even true within mathematics*. The concept of risk aversion treats risk as "noise", as though it is something that only exists to be ignored, but if the common place use of biased MLE's is an indication of anything, it is that *ignoring uncertainty* is the irrational choice.

Trust is another concept that is related to uncertainty communication, but fact that it is often front and center in the discussion has a somewhat ominous subtext. If the purpose of displaying uncertainty information is to appropriately hedge a signal with noise, then it should be assumed that trust is only related to uncertainty communication through increased transparency and honesty. It does seem to be the case that uncertainty communication increases trust because it is a proxy for *(add citations)*. Unfortunately a large proportion of the literature discusses trust as something that is directly related to uncertainty visualisation rather than as an observable product. In viewing uncertainty communication as directly related to trust, not related through the proxy of transparency, several unobserved variables are conflated. The amount of uncertainty that is present in the information, whether or not the information is trustworthy or makes sense, prior beliefs of the participants, and the trustworthiness of the source are a few examples of variables that are conflated when authors directly consider trust to be of direct interest. Similarly to the issues in risk-aversion directly considering trust, and not transparency, as the metric of importance in uncertainty communication leads to a questionable subtext that argues against transparency. Being concerned about audiences perception of trust, without first establishing if what we are communicating is *trustworthy*, leads to the implication that details that result in information not being considered trustworthy should be avoided.

Authors that imply that a decrease in trust or an increase in ambiguity or risk aversion are metrics of importance in of themselves do not understand the purpose of uncertainty communication as methods to increase transparency. Science communication should be primarily concerned with accuracy, setting trust and risk-aversion as the variables of interest implicitly encourages statisticians to set trust and risk-aversion as the primary goals of communication. The issue of trust being divorced from trustworthiness has been commented on by other authors [@ONeill2018], however the issue still persists in the uncertainty visualisation literature [@Zhao2023]. Additionally, it seems that risk-aversion being divorced from the value of risk has gone unnoticed.

While this is a focus on uncertainty visualisation, these issues in uncertainty communication need to be acknowledged because they bleed into the uncertainty visualisation literature. Uncertainty visualisation are compared on their ability to communicate explicit risk, elicit trust, and prevent risk or ambiguity avoidance, continuing these issues in uncertainty communication at large. While uncertainty visualisation can make it easier to communicate the complicated details of uncertainty information, it carries some challenges that are unique to the visualisaition, specifically with respect to the "signal supression" concept. In simple estimates or verbal communication, the signal is often easy to identify because it is what we are explicitly saying. Visualisations are used in both data exploration and communication. This means what exactly is a *signal* in any particular visualisation is hard to identify, since we often let the visualisation *tell us* what the signal is. Additionally, you cannot add noise to *every single possible* signal one might take from a visualisation. Two people looking at the same visualisation might, just by chance, develop two entirely different insights. These unique and facinating challenges that are faced by viewing uncertainty visualisaiton through the lense of "signal supression" have been almost completely untouched by the literature.

# 4 LITERATURE SUMMARY
- methods used in uncertainty visualisation evaluation remain ad hoc. studies approach issue from a usability perspective instead of asking WHY representations DO or DO NOT work. Studies do not follow any methodology commonly agreed upon. @Kinkeldey2014
- Lack of formalisation and eigour in empirical methdos is an issue that is much broader than uncertainty visualisation (extends to info-vis and related domains). Therefore this paper is a starting point, not a definitive summary @Kinkeldey2014
- The question in these papers is often "does method a work better than method B" which is an engineering approach, where the toal is to improve a particular product rather than create general principals @Kinkeldey2014
- Authors then try to generalise beyond the specific constraints of the test even though the framing does not allow for it @Kinkeldey2014
- Alternatively, we should ask how/why does method A work better than method B, which grounds work in perceptual and cognative theory and provides a framwork @Kinkeldey2014
